A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}

### 2011-02-15 introduction.tex
lloid_paper intro: reworked introduction quite a bit
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@625 ccdd05d5-56c1-461a-bf98-85f6876523b4

% I would like to say:
%  - Low mass inspirals are the most promising source of gravitational radiation for advanced LIGO
% - Mention S6 and VSR2, online running, EM followup
% - Observational applications of near-realtime detection
% - Point out challenges of advanced detectors: more templates, longer templates
% - Advanced LIGO timetable
% - Einstein telescope
% - Layout of paper: description of method, procedure followed

\begin{comment}

However, the computational requirements of conventional detection strategies --- especially those that are well suited to low latency applications --- will be greatly increased by this upgrade for two reasons.  First, advanced detectors will have much improved low-frequency sensitivity, so chirp signals from coalescing binaries will remain in the detectors' band for longer.  This will demand that matched filter based searches use longer templates, requiring more memory and more cycles to process.  Second, background noise for advanced detectors will be more spectrally uniform, so it will be possible to resolve the intrinsic parameters of a source that determine the time-frequency structure of the gravitational wave signal with much greater accuracy.  This will require that matched filter bank based searches employ more matched filters, again demanding more memory and more cycles.

\end{comment}

\begin{comment}

\editorial{This paragraph has too many citations.} Modern gravitational-wave observatories are likely to deliver on the promise 
that their name implies by observing multiple gravitational-wave signals from
compact binary coalescence within the coming decade \cite{kopparapu2008}. 
There is now a world-wide network of gravitational-wave observatories 
including the laser interferometers
\textsc{ligo}, Virgo, \textsc{geo}, and \textsc{tama}, which should assure that coincident detections 
will be possible among at least three separate sites lending confidence
to any detections~\cite{Abbott:2005qm, beauville2008, beauville2006, blackburn2005, LIGOS1instpaper}.  \editorial{This is a lot of citations just to say that the 4 detectors exist.  Maybe if they were scattered throughout the paragraph?}
However, having an electromagnetic or neutrino counterpart to a 
gravitational-wave detection would not only increase the confidence in 
the detection but would also greatly improve the astrophysical information
available from the event.

\end{comment}

In a series of detector upgrades that are now under way, \textsc{ligo} and Virgo will become the first advanced gravitational wave detectors, gaining a tenfold improvement in amplitude sensitivity \citeneeded{} and a corresponding thousandfold increase in observable volume in the local universe.  It is anticipated that these detectors will be able to sense as many as 40 events per year \cite{Abadie:2010p10836}.

As the gravitational wave detection horizon pushes outward, the ability to detect signals in near realtime will become valuable.  Having an electromagnetic or neutrino counterpart to a 
gravitational-wave detection would not only increase the confidence in 
the detection but will also greatly improve the astrophysical information
available from the event.  Most models that predict simultaneous
gravitational-wave and electromagnetic observations also predict that the
peak amplitude of electromagetic radiation will occur soon after 
gravitational-wave emission~\cite{sylvestre2003}.  Thus in order to maximize
the chance of a successful electromagnetic followup the latency of 
gravitational-wave signal analysis must be made minimal.

\editorial{This is old news, and doesn't seem relevant to me: EM-directed gravitational wave searches don't benefit from low-latency detection pipelines.} Work has already commenced to trigger gravitational-wave searches from 
electromagnetic observations~\cite{triggeredsearches2008}. \textsc{ligo} observations
ruled out the possibility of GRB070201 originating from the merger of a
\textsc{nsns} or \textsc{nsbh} system in the Andromeda galaxy~\cite{GRB070201}.

In \oldstylenums{2010}, \textsc{ligo} and Virgo completed a period of joint data taking during which several all-sky detection pipelines operated in a low-latency configuration \citeneeded.  Promising detection candidates were promptly sent for electromagnetic followup \citeneeded{} to several telescopes including Swift, \textsc{rotse}, \textsc{tarot}, and Zadko.

This question has been investigated and it seems  at least
possible that gravitational-wave detections could provide a region of the
sky to prompt electromagnetic observation followups~\cite{sylvestre2003}.
\editorial{Cite an in-preparation paper of Larry's?} Much work is currently underway in providing the best possible source 
localization from gravitational-wave detector networks~\cite{markowitz2008, raymond2008, cavalier2006}.  Collaborations are currently forming to provide infrastructure for the gravitational-wave
Astronomers to provide targets of opportunity for electromagnetic 
astronomers~\cite{kanner2008}.  Most models that predict coincident
gravitational-wave and electromagnetic observations also predict that the
peak electromagetic fluence will occur soon after 
gravitational-wave emission~\cite{sylvestre2003}.  Thus in order to maximize
the chance of a successful electromagnetic followup the latency of 
gravitational-wave signal analysis must be made minimal.

The parameter space of compact binary coalescence signals is large~\cite{Owen:1995tm, Owen:1998dk}.  It is a
computationally burdensome task to analyze these signals with even moderate
latency~\cite{Abbott:2007xi}.  
This work will describe how to exploit degeneracy in the signal
parameter space to answer more quickly whether or not a gravitational-wave 
is present. Specifically we will explore using the singular value decomposition (\textsc{svd}) to reduce the
effective number of filters required to search the data.  We note that others
have applied the use of \textsc{svd} to gravitational wave data analysis to analyze
optimal gravitational-wave burst detection~\cite{bradyraymajumder2004, heng2008} and coherent networks 
of detectors~\cite{wen2008} .

>>>>

% Outline:
%  - Low-mass inspirals are the most promising source of gravitational radiation for advanced \textsc{ligo}
% - Motivation for high throughput
%   - Point out challenges of advanced detectors: more templates, longer templates
% - Motivation for low latency
%   - Observational applications of near-realtime detection
%   - Mention S6 and VSR2, online running, EM followup
% - Layout of paper: description of method, procedure followed

Low-mass inspiraling binaries are the most promising sources of gravitational
radiation for Advanced \textsc{ligo}. Observation demonstrates that they must
exist in the local Universe in some abundance~\cite{Abadie:2010p10836}.
[science] In the
final stages of inspiral, the neutron star is tidally disrupted, providing the
matter required to produce bright electromagnetic
counterparts~\cite{shibata:2007}. Conveniently, the gravitational waveforms are
understood and implemented to the level required by Advanced \textsc{ligo} to
detect~\cite{BuonannoIyerOchsnerYiSathya2009}, allowing the sensitive technique
of matched filtering.

The parameter space of compact binary coalescence signals is
large~\cite{Owen:1995tm, Owen:1998dk} leading to large filter banks and
significant computational cost. Following the series of detector upgrades that
are now under way, \textsc{ligo} and Virgo will attain their Advanced
configurations, gaining a tenfold improvement in amplitude sensitivity over
Initial configurations~\cite{advLIGOrefdesign,advVirgorefdesign} and a
corresponding thousandfold increase in observable volume in the local Universe.
The computational requirements of conventional detection strategies will be
greatly increased by this upgrade for two reasons. First, advanced detectors
will have much improved low-frequency sensitivity, so chirp signals from
coalescing binaries will remain in the detectors' band for a significantly
longer time. This will demand that matched filter based searches use longer
templates, requiring more memory and more cycles to process. Second, background
noise for advanced detectors will be more spectrally uniform, so it will be
possible to resolve the intrinsic parameters of a source that determine the
time-frequency structure of the gravitational wave signal with much greater
accuracy. This will require that matched filter bank based searches employ
more matched filters, again demanding more memory and more cycles.

Beyond raw throughput, the ability to detect signals in near realtime will
become increasingly valuable as the gravitational-wave detection horizon pushes
outward.  Having an electromagnetic or neutrino counterpart to a
gravitational-wave detection would not only increase the confidence in the
detection but will also greatly increase the astrophysical information
available from the event. Most models that predict simultaneous
gravitational-wave and electromagnetic observations also predict that the peak
amplitude of electromagetic radiation will occur soon after gravitational-wave
emission~\cite{sylvestre2003}. Thus in order to maximize the chance of a
successful electromagnetic followup the latency of gravitational-wave signal
analysis must be made minimal.

In \oldstylenums{2010}, \textsc{ligo} and Virgo completed S6/VSR3,
a short period of
joint data taking during which several all-sky detection pipelines operated in
a low-latency configuration \citeneeded. A few candidates of moderate
significance were promptly sent for electromagnetic followup \citeneeded{} to
several telescopes including Swift, \textsc{rotse}, \textsc{tarot}, and Zadko.
They achieved latencies of BLAH, but required human vetting of each candidate
before alerting the telescope partners. As confidence in the infrastructure
increases, humans can be removed from the loop, setting a demonstrated previous
best latency at BLAH.

This work will describe how to exploit degeneracy in the signal parameter space
to answer more quickly and economically whether or not a gravitational wave is
present.
\editorial{nvf: I use some rather different words here to summarize the
techniques. I don't want to simply repeat words from later, but I don't want
to jar a reader with changing jargon either. Feel free to edit.}
Specifically, we will exploit the time-frequency structure of chirp
signals to downsample different parts of the waveform, use the singular value
decomposition (\textsc{svd}) factors to identify the redundancy of the bank and
reduce the effective number of filters required to search the data, and use
matched subspace filters built from \textsc{svd} to identify when computation
can be avoided.  We note that others have applied the use of \textsc{svd} to
gravitational wave data analysis to analyze optimal gravitational-wave burst
detection~\cite{bradyraymajumder2004, heng2008} and coherent networks of
detectors~\cite{wen2008}.
